---
title: 'Detailed `edgeR` Workflow on Prostate Cancer RNASeq Analysis Between African Americans and Causasian Americans'
author: "Flemming Wu"
---

```{r setup, include=FALSE}
if (Sys.info()["sysname"] == "Darwin") {setwd("~/Desktop/velvilab/P_Ca_Analysis")}
if (Sys.info()["sysname"] == "Windows") {setwd("C:/Users/flemm/velvilab/P_Ca_Analysis_012723")}
```

```{r load libs, message=F}
library(readxl)
library(openxlsx)
library(magrittr)
library(purrr)
library(tidyverse)
library(RUVSeq)
library(RColorBrewer)
library(DESeq2)
library(EnsDb.Hsapiens.v86)
library(EnhancedVolcano)
library(clusterProfiler)
library(org.Hs.eg.db)
library(ggpubr)
library(msigdbr)
library(enrichplot)
library(pathview)
```

```{r read in data, message=F}
metadata <- read_xlsx("./data/RNAseq_Samples_012723.xlsx")
data <- read_xlsx("./data/Data_012723.xlsx")
```

```{r prepare data}
# rows containing Ensembl id are the same, set to rownames and remove columns, keeping only read count columns
data <- column_to_rownames(data, var = "Geneid...1") %>%
  dplyr::select(-starts_with("Geneid"))

# change sample names to make it shorter for downstream workflow
new.names <- metadata %>%
  mutate(KGP_Sample_ID = stringr::str_extract(KGP_Sample_ID, "00[:digit:]{2}") %>% 
           stringr::str_replace("00[0]?", "RNA")) %>%
  mutate(`Tissue Type` = ifelse(`Tissue Type` == "Malignant", "M", "N"), 
                Race = ifelse(Race == "Black", "B", "W")) %>%
  tidyr::unite(id, c("Tissue Type", "Race", "MGC_Sample_ID")) %>%
  arrange(as.integer(str_extract(KGP_Sample_ID, "[:digit:]{1,2}"))) %>%
  dplyr::select(id) %>%
  as.list()

names(data) <- new.names$id

head(data)
```

```{r split by comparisons}
## split data for all comparisons

# Nonnmalignant black vs malignant black
mb_nb <- data %>% select_if(grepl("M_B", names(.)) | grepl("N_B", names(.)))
# Walignant white vs nonmalignant black
mw_nb <- data %>% select_if(grepl("M_W", names(.)) | grepl("N_B", names(.)))
# Nonmalignant black vs nonmalignant white
nb_nw <- data %>% select_if(grepl("N_B", names(.)) | grepl("N_W", names(.)))
# Malignant black vs malignant white
mb_mw <- data %>% select_if(grepl("M_B", names(.)) | grepl("M_W", names(.)))
# Malignant black vs nonmalignant white
mb_nw <- data %>% select_if(grepl("M_B", names(.)) | grepl("N_W", names(.)))
# Malignant white vs nonmalignant white
mw_nw <- data %>% select_if(grepl("M_W", names(.)) | grepl("N_W", names(.)))

## combine into list of datasets
all.dfs <- list("mb_nb" = mb_nb, "mw_nb" = mw_nb, "nb_nw" = nb_nw, "mb_mw" = mb_mw, "mb_nw" = mb_nw, "mw_nw" = mw_nw)
```

### Remove unwanted variation with `RUVg`

First, filter out rows that don't meet the requirements of: more than 5 reads in at least 3 samples from each comparison

```{r filter lowly expressed genes}
all.dfs.filt <- lapply(all.dfs, function(df) df[apply(df, 1, function(x) length(x[x>5])>=3), ])
```

```{r}
lapply(all.dfs.filt, head)
```

Create SeqExpressionSet objects from data frames

```{r}
make_sets <- function(df) {
  x <- names(df) %>% str_extract("[MN]_[WB]") %>% as.factor()
  set <- newSeqExpressionSet(as.matrix(df), phenoData = data.frame(x, row.names = colnames(df)))
  return(set)
}

all.sets <- lapply(all.dfs.filt, function(x) make_sets(x))
```

```{r}
str(all.sets$mb_nb)
```

#### EDA

View PCA and RLE plots of the samples prior to normalization

```{r}
plot_sets <- function(set) {
  x <- set$x
  colors <- brewer.pal(3, "Set1")
  plotRLE(set, outline = FALSE, ylim = c(-4,4), col = colors[x])
  plotPCA(set, col = colors[x], cex = 0.7)
}

lapply(all.sets, function(set) plot_sets(set)) ## check plots
```

#### Run between lane (upper quartile) normalization and re-visualize

```{r between lane normalization}
all.sets.norm <- lapply(all.sets, function(set) betweenLaneNormalization(set, which = "upper"))

lapply(all.sets.norm, function(set) plot_sets(set))
```


```{r view non-normalized count matrix}
head(counts(all.sets.norm$mb_nb))
```

```{r view normmalized count matrix}
head(normCounts(all.sets.norm$mb_nb))
```

### Create empirical negative control genes and Run DE analysis with `edgeR`

Since no genes are known a priori to be influenced by covariates of interest, you can run a first-pass differential expression analysis and the least significantly DE genes found from the analysis can be used as negative controls for the second-pass differential expression analysis.

```{r estimate negative controls}
## write function to update SeqExpressionSet objects to have negative controls
est_neg_ctrl <- function(set) {
  x <- set$x
  design <- model.matrix(~x, data = pData(set))
  y <- DGEList(counts = counts(set), group = x)
  y <- calcNormFactors(y, method = "upperquartile")
  y <- estimateGLMCommonDisp(y, design)
  y <- estimateGLMTagwiseDisp(y, design)

  fit <- glmFit(y, design)
  lrt <- glmLRT(fit, coef = 2)

  top <- topTags(lrt, n = nrow(set))$table
  empirical <- rownames(set)[which(!(rownames(set) %in% rownames(top)[1:5000]))]
  
  # run RUVg
  newset <- RUVg(set, empirical, k = 1)
  return(newset)
}

all.sets.norm.nc <- lapply(all.sets.norm, est_neg_ctrl)
```

```{r}
pData(all.sets.norm.nc$mb_nb)
```













































